---
title: "erm-1 RNAi Treatments — Merged 2-cell 4-cell Analysis"
author: "Sam Zavislan-Pullaro"
date: "2026-07-26"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Load libraries

```{r setup}
library(tidyverse)
library(data.table)
library(rstatix)
library(ggrepel)
library(gridExtra)
```

---

## Import data

Note: channel 1 = set-3 mRNA | channel 2 = erm-1 mRNA

L4440 control files are loaded once and shared across all treatment comparisons.

```{r import}

col_times <- paste(rep(1:333), sep = "")

# L4440 empty vector 2-cell stage 

control_1 <- read.table("../01_input/2-cell/2026-6-22_1458_datafile_for_230713_L4440_RNAi_rep1.txt",
                        header = FALSE, sep = "\t")
control_2 <- read.table("../01_input/2-cell/2026-6-22_153_datafile_for_230828_L4440_RNAi_rep2.txt",
                        header = FALSE, sep = "\t")
control_3 <- read.table("../01_input/2-cell/2026-6-22_156_datafile_for_240122_LP306_L4440-rep3.txt",
                        header = FALSE, sep = "\t")

colnames(control_1) <- c("file", "channel", col_times)
colnames(control_2) <- c("file", "channel", col_times)
colnames(control_3) <- c("file", "channel", col_times)

# L4440 empty vector 4-cell stage (ABa-ABp)

control_1_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-11_1252_datafile_for_230713_L4440_RNAi_rep1.txt",
                        header = FALSE, sep = "\t")
control_2_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-11_133_datafile_for_230828_L4440_RNAi_rep2.txt",
   header = FALSE, sep = "\t")
control_3_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-11_1337_datafile_for_240122_LP306_L4440-rep3.txt",
                        header = FALSE, sep = "\t")
control_4_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-11_1324_datafile_for_231107_LP306_L4440.txt",
                        header = FALSE, sep = "\t")

colnames(control_1_ABaABp) <- c("file", "channel", col_times)
colnames(control_2_ABaABp) <- c("file", "channel", col_times)
colnames(control_3_ABaABp) <- c("file", "channel", col_times)
colnames(control_4_ABaABp) <- c("file", "channel", col_times)

# bmk-1 RNAi KD 2-cell stage 

bmk1_1 <- read.table("../01_input/2-cell/2026-6-22_1528_datafile_for_220719_LP306_bmk-1_RNAi.txt",
                     header = FALSE, sep = "\t")
bmk1_2 <- read.table("../01_input/2-cell/2026-6-22_1536_datafile_for_230408_LP306_BMK-1_rep1.txt",
                     header = FALSE, sep = "\t")
bmk1_3 <- read.table("../01_input/2-cell/2026-6-22_1544_datafile_for_231107_LP306_BMK-1_rep2.txt",
                     header = FALSE, sep = "\t")
bmk1_4 <- read.table("../01_input/2-cell/2026-6-22_1550_datafile_for_231113_LP306_BMK-1_rep3.txt",
                     header = FALSE, sep = "\t")

colnames(bmk1_1) <- c("file", "channel", col_times)
colnames(bmk1_2) <- c("file", "channel", col_times)
colnames(bmk1_3) <- c("file", "channel", col_times)
colnames(bmk1_4) <- c("file", "channel", col_times)

# bmk-1 RNAi KD 4-cell stage (ABa-ABp)

bmk1_1_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-14_126_datafile_for_220719_LP306_bmk-1_RNAi.txt",
                     header = FALSE, sep = "\t")
bmk1_2_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-14_1216_datafile_for_230408_LP306_BMK-1_rep1.txt",
                     header = FALSE, sep = "\t")
bmk1_3_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-14_1222_datafile_for_231107_LP306_BMK-1_rep2.txt",
                     header = FALSE, sep = "\t")
bmk1_4_ABaABp <- read.table("../01_input/4-cell_ABaABp/2026-7-14_1225_datafile_for_231113_LP306_BMK-1_rep3.txt",
                     header = FALSE, sep = "\t")

colnames(bmk1_1_ABaABp) <- c("file", "channel", col_times)
colnames(bmk1_2_ABaABp) <- c("file", "channel", col_times)
colnames(bmk1_3_ABaABp) <- c("file", "channel", col_times)
colnames(bmk1_4_ABaABp) <- c("file", "channel", col_times)

```

---

## Create annotation columns and add timepoints

Timepoints are treated as positions along line scan

```{r annotate}

# Helper function: pivot longer + annotate 
annotate_df <- function(df) {
  df[2:nrow(df), 1:335] %>%
    separate_wider_delim(file, delim = "_",
                         names = c("date", "strain", "treatment", "embryoID", NA)) %>%
    pivot_longer(cols = `1`:`333`, names_to = "xpoint", values_to = "intensity") %>%
    mutate(unique_id = paste(date, strain, treatment, embryoID, channel, sep = "_"))
}

# Helper: add timepoints from the header row of any loaded table. Timepoints used as proxy for position
addInTimepoints <- function(tibble, ref_table) {
  timepoints <- as.vector(unlist(ref_table[1, 3:335]))
  repnumb    <- nrow(tibble) / 333
  tibble::add_column(tibble, timepoints = rep(timepoints, repnumb), .after = "xpoint")
}

# Controls 2-cell AB-P1
# Note: boundary tag added to every set below. This isn't strictly needed for
# the 2-cell sets (different embryos, no ID collision risk) but is included
# for consistency/bookkeeping. It IS required for the 4-cell sets, since
# ABaABp and ABaEMS come from the SAME embryo and would otherwise share an
# identical date/strain/treatment/embryoID/channel key.

ctrl1_pt <- addInTimepoints(annotate_df(control_1), control_1) %>% mutate(boundary = "ABP1") # control_1 used as a reference to set position
ctrl2_pt <- addInTimepoints(annotate_df(control_2), control_1) %>% mutate(boundary = "ABP1")
ctrl3_pt <- addInTimepoints(annotate_df(control_3), control_1) %>% mutate(boundary = "ABP1")

# Controls 4-cell ABa-ABp

control_1_ABaABp_pt <- addInTimepoints(annotate_df(control_1_ABaABp), control_1) %>% mutate(boundary = "ABaABp")
control_2_ABaABp_pt <- addInTimepoints(annotate_df(control_2_ABaABp), control_1) %>% mutate(boundary = "ABaABp")
control_3_ABaABp_pt <- addInTimepoints(annotate_df(control_3_ABaABp), control_1) %>% mutate(boundary = "ABaABp")
control_4_ABaABp_pt <- addInTimepoints(annotate_df(control_4_ABaABp), control_1) %>% mutate(boundary = "ABaABp")

# bmk-1 2-cell AB-P1
bmk1_1_pt <- addInTimepoints(annotate_df(bmk1_1), control_1) %>% mutate(boundary = "ABP1")
bmk1_2_pt <- addInTimepoints(annotate_df(bmk1_2), control_1) %>% mutate(boundary = "ABP1")
bmk1_3_pt <- addInTimepoints(annotate_df(bmk1_3), control_1) %>% mutate(boundary = "ABP1")
bmk1_4_pt <- addInTimepoints(annotate_df(bmk1_4), control_1) %>% mutate(boundary = "ABP1")

# bmk-1 4-cell ABa-ABp

bmk1_1_ABaABp_pt <- addInTimepoints(annotate_df(bmk1_1_ABaABp), control_1) %>% mutate(boundary = "ABaABp")
bmk1_2_ABaABp_pt <- addInTimepoints(annotate_df(bmk1_2_ABaABp), control_1) %>% mutate(boundary = "ABaABp")
bmk1_3_ABaABp_pt <- addInTimepoints(annotate_df(bmk1_3_ABaABp), control_1) %>% mutate(boundary = "ABaABp")
bmk1_4_ABaABp_pt <- addInTimepoints(annotate_df(bmk1_4_ABaABp), control_1) %>% mutate(boundary = "ABaABp")

```

---

## Merge all data and apply fixed coordinate alignment

L4440 controls are included once (shared reference across all experiments).
Peak alignment is skipped: every embryo receives the same fixed coordinate,
recentered so xpoint 1:333 maps to aligned_row -166:166.

```{r merge_align}

all_data <- rbind(
  ctrl1_pt, ctrl2_pt, ctrl3_pt, 
  control_1_ABaABp_pt, control_2_ABaABp_pt, control_3_ABaABp_pt, control_4_ABaABp_pt,
  bmk1_1_pt, bmk1_2_pt, bmk1_3_pt, bmk1_4_pt, 
  bmk1_1_ABaABp_pt, bmk1_2_ABaABp_pt, bmk1_3_ABaABp_pt, bmk1_4_ABaABp_pt
)

dt <- as.data.table(all_data)
dt$xpoint <- as.integer(dt$xpoint)
dt[, aligned_row := xpoint - 167]

total_align_long <- dt %>%
  select(strain, aligned_row, unique_id, intensity, boundary)

cat("Total rows:", nrow(total_align_long), "\n")
cat("Unique embryo IDs:", n_distinct(total_align_long$unique_id), "\n")

```

---

## Normalize

Each embryo is normalized to its own mean intensity (per embryo per channel).

```{r normalize}

norm_total <- total_align_long %>%
  separate_wider_delim(unique_id, delim = "_",
                       names = c("date", NA, "treatment", "embryoID", "channel")) %>%
  group_by(date, embryoID, channel, treatment, boundary) %>%
  mutate(normalized_intensity = intensity / mean(intensity, na.rm = TRUE)) %>%
  ungroup()

treatment_order <- c("L4440", "bmk-1")
norm_total$treatment <- factor(norm_total$treatment, levels = treatment_order)

# Label each embryo as control or RNAi for coloring
norm_total <- norm_total %>%
  mutate(treatment_type = ifelse(treatment == "L4440", "L4440", "RNAi"))

table(norm_total$channel, norm_total$treatment)

```

---

## Color palettes

```{r palettes}

# Per-treatment palette for line scans
treatment_colors <- c(
  "L4440"   = "gray50",
  "bmk-1"   = "#17becf"
)

# Blue/grey scheme matching individual scripts (for boxplots)
type_colors <- c("L4440" = "darkgray", "RNAi" = "lightblue")

channel_labels <- c("ch1" = "set-3 mRNA", "ch2" = "erm-1 mRNA")

```

---

## Section 1: Individual normalized line scans

All embryo-level line scans, faceted by channel and treatment.

```{r linescans_individual, fig.height = 12, fig.width = 14}

p_linescan_individual <- ggplot(norm_total,
       aes(x = aligned_row, y = normalized_intensity,
           group = interaction(date, embryoID),
           color = treatment)) +
  geom_line(alpha = 0.4, linewidth = 0.3) +
  scale_color_manual(values = treatment_colors) +
  facet_grid(channel ~ treatment,
             labeller = labeller(channel = channel_labels)) +
  labs(
    title = "Individual normalized line scans — merged 2-cell and 4-cell",
    x     = "Position (aligned_row)",
    y     = "Normalized intensity",
    color = "Treatment"
  ) +
  guides(color = "none") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

p_linescan_individual

```

---

## Section 2: Mean +- SD line scans

### 2a. All treatments overlaid, one panel per channel

```{r linescans_mean_overlaid, fig.height = 6, fig.width = 12}

mean_signal <- norm_total %>%
  group_by(aligned_row, treatment, channel) %>%
  summarize(
    mean_signal = mean(normalized_intensity, na.rm = TRUE),
    sd_signal   = sd(normalized_intensity,   na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  mutate(
    ymeanmax = mean_signal + sd_signal,
    ymeanmin = mean_signal - sd_signal
  )

ggplot(mean_signal,
       aes(x = aligned_row, y = mean_signal,
           color = treatment, fill = treatment,
           group = treatment)) +
  geom_ribbon(aes(ymin = ymeanmin, ymax = ymeanmax), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = treatment_colors) +
  scale_fill_manual(values  = treatment_colors) +
  facet_wrap(~ channel, labeller = labeller(channel = channel_labels)) +
  labs(
    title  = "Mean ± SD normalized line scans — all treatments",
    x      = "Position (aligned_row)",
    y      = "Mean normalized intensity",
    color  = "Treatment",
    fill   = "Treatment"
  ) +
  theme_bw(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "right")

```

### 2b. L4440 vs each RNAi treatment, faceted grid

```{r linescans_mean_faceted, fig.height = 10, fig.width = 14}

ggplot(mean_signal,
       aes(x = aligned_row, y = mean_signal,
           group = treatment)) +
  geom_ribbon(aes(ymin = ymeanmin, ymax = ymeanmax,
                  fill = treatment), alpha = 0.2, color = NA) +
  geom_line(aes(color = treatment), linewidth = 0.8) +
  scale_color_manual(values = treatment_colors) +
  scale_fill_manual(values  = treatment_colors) +
  facet_grid(channel ~ treatment,
             labeller = labeller(channel = channel_labels)) +
  labs(
    title = "Mean +- SD line scans, per treatment, per channel",
    x     = "Position",
    y     = "Mean normalized intensity"
  ) +
  guides(color = "none", fill = "none") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

```

---

## Section 3: log2 fold change enrichment

Ratio of normalized intensity at position 1 versus the mean of positions
-100 and +100. Higher values indicate more enriched localization signal
(consistent with membrane localization).

```{r}

peaks_and_valleys <- norm_total %>%
  filter(aligned_row %in% c(-100, 1, 100))

nest_pandv <- peaks_and_valleys %>%
  group_by(date, treatment, embryoID, channel, boundary) %>%
  nest()

my_calc2 <- function(df) {
  df$normalized_intensity[df$aligned_row == 1] /
    mean(df$normalized_intensity[df$aligned_row %in% c(-100, 100)])
}

foldChange_calc <- nest_pandv %>%
  mutate(fc_enrich = map_dbl(data, my_calc2)) %>%
  mutate(
    log2_fc        = log(fc_enrich, base = 2),
    treatment_type = ifelse(treatment == "L4440", "L4440", "RNAi"),
    treatment      = factor(treatment, levels = treatment_order)
  )

summary(foldChange_calc$log2_fc)

```

### 3a. Merged boxplot, x-axis by treatment, faceted by channel

```{r boxplot_merged, fig.height = 6, fig.width = 12}

l4440_ch2_ref <- foldChange_calc %>%
  ungroup() %>%
  filter(treatment == "L4440", channel == "ch2") %>%
  summarize(med = median(log2_fc, na.rm = TRUE)) %>%
  mutate(channel = "ch2")

p_merged <- ggplot(foldChange_calc,
                   aes(x = treatment,
                       y = log2_fc,
                       fill = treatment_type)) +
  geom_boxplot(outlier.shape = NA, width = 0.6) +
  geom_jitter(aes(color = treatment_type),
              position = position_jitterdodge(jitter.width = 0.5),
              size = 1.5, alpha = 0.7) +
  scale_fill_manual(
    values = type_colors,
    labels = c("L4440" = "L4440 control", "RNAi" = "RNAi treatment")
  ) +
  scale_color_manual(
    values = c("L4440" = "gray30", "RNAi" = "#1a6fa0"),
    guide  = "none"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(data = l4440_ch2_ref, aes(yintercept = med),
             linetype = "dashed", color = "red", linewidth = 0.6) +
  scale_y_continuous(limits = c(-0.5, 1.5)) +
  facet_wrap(~ channel, labeller = labeller(channel = channel_labels)) +
  labs(
    title = "log2 enrichment at position 0 vs ±100",
    x     = "Treatment",
    y     = "log2 (intensity at 0 / mean intensity at ±100)",
    fill  = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1, face = "italic"),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )

p_merged

```

### 3b. Merged boxplot, x-axis by treatment, violin plots

```{r violinplot_merged, fig.height = 6, fig.width = 12}

l4440_ch2_ref <- foldChange_calc %>%
  ungroup() %>%
  filter(treatment == "L4440", channel == "ch2") %>%
  summarize(med = median(log2_fc, na.rm = TRUE)) %>%
  mutate(channel = "ch2")

violinplot_merged <- ggplot(foldChange_calc,
                   aes(x = treatment,
                       y = log2_fc,
                       fill = treatment_type)) +
  geom_violin(width = 0.6, alpha = 0.8, color = NA) +
  geom_boxplot(width = 0.12, fill = "white", alpha = 0.7,
               outlier.shape = NA, position = position_dodge(width = 0.6)) +
  geom_jitter(aes(color = treatment_type),
              position = position_jitterdodge(jitter.width = 0.5),
              size = 1.2, alpha = 0.5) +
  scale_fill_manual(
    values = type_colors,
    labels = c("L4440" = "L4440 control", "RNAi" = "RNAi treatment")
  ) +
  scale_color_manual(
    values = c("L4440" = "gray30", "RNAi" = "#1a6fa0"),
    guide  = "none"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_y_continuous(limits = c(-0.5, 1.5)) +
  facet_wrap(~ channel, labeller = labeller(channel = channel_labels)) +
  labs(
    title = "log2 enrichment at position 0 vs \u00b1100",
    x     = "Treatment",
    y     = "log2 (intensity at 0 / mean intensity at \u00b1100)",
    fill  = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1, face = "italic"),
    strip.text      = element_text(face = "bold"),
    legend.position = "top",
    plot.title      = element_text(face = "bold", hjust = 0.5)
  )
violinplot_merged

```
---

## Summary statistics

```{r summary_stats}

summary_stats_table <- foldChange_calc %>%
  group_by(treatment, channel) %>%
  summarise(
    n          = n(),
    min_log2fc = round(min(log2_fc,    na.rm = TRUE), 3),
    median_log2fc = round(median(log2_fc, na.rm = TRUE), 3),
    mean_log2fc   = round(mean(log2_fc,   na.rm = TRUE), 3),
    max_log2fc = round(max(log2_fc,    na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(channel, treatment)

print(summary_stats_table, n = Inf)

```

---

## Statistics - Pairwise Wilcoxon vs L4440

```{r pairwise_wilcox}

# Pairwise Wilcoxon vs L4440 reference (per channel)
pairwise_stats <- foldChange_calc %>%
  group_by(channel) %>%
  wilcox_test(log2_fc ~ treatment, ref.group = "L4440") %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

pairwise_stats

```

---

## Session info

```{r sessionInfo}
sessionInfo()
```

---

## Export plots and stats

```{r export_outputs}

today      <- format(Sys.Date(), "%y%m%d")
output_dir <- "../03_output/02_output_2-cell_4_cell_ABa-ABp_merged"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(file.path(output_dir, paste0(today, "_boxplot_merged.svg")), plot = p_merged,
       width = 12, height = 7)
ggsave(file.path(output_dir, paste0(today, "_linescan_individual.svg")), plot = p_linescan_individual,
       width = 14, height = 12)

write_csv(summary_stats_table, file.path(output_dir, paste0(today, "_summary_stats.csv")))
write_csv(pairwise_stats, file.path(output_dir, paste0(today, "_pairwise_wilcoxon_stats.csv")))

cat("Exported to:", normalizePath(output_dir), "\n")

```
